MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2022-02-14 main/MSstatsPTM_Main.tex
Abstract+intro

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges which are statistical in nature. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimating relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The method measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove bias from changes in the overall protein. We implemented this model in the free and open-source R package $MSstatsPTM$. The approach is evaluated on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with varying acquisition methods. Finally, the approach is compared against several existing methods and is shown to outperform all other methods.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe} \cite{Cohen}. Mass spectrometry-based proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}. Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}. Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward\cite{Olsen:2013}. Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.

There are two main biological questions that may be relevant in PTM analysis. First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers.\cite{Petushkova_2017}. In this case the researcher may not necessarily want to correct for confounding with the total protein. The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups. Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions. This is an important question for thinking about biological and physiological models and very short timescale experiments. In this case, we are interested in the relative abundance change solely from the modification and correcting for confounding with the total protein is relevant. These two questions are both important areas of study and require different statistical methods to correctly address. In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods that account for confounding with the total protein.

We propose a general statistical approach, which explicitly characterizes the variations and confounding factors present in bottom-up PTM experiments. The proposed approach is aimed at the detection of quantitative changes in PTMs between conditions utilizing procedures developed for summarization of LC-MS data, quantitative characterization of site-specific PTMs, and adjustment with respect to protein abundance. Quantitative analyses of PTMs often involve comparisons between multiple inter-related conditions of the same biological system. The general statistical framework underlying the proposed approach allows for analyzing experiments with complex designs and different acquisition methods.

The proposed approach was implemented using the R coding language and evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was compared against the commonly applied t-test and Limma methods\cite{Ritchie_15a}. The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate. Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014} \cite{Huang:2020}.

@@

This section describes the datasets and experiments used to evaluate the proposed method. The datasets were chosen to provide a performance benchmark of the proposed approach against existing methods when ground truth is known, as well as represent a variety of experimental designs and acquisition methods where the approach can be applied. To benchmark the approach, computer simulations and a spike-in experiment were used. The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment where ground truth is known. Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and organism. These experiments were chosen to show the proposed method's applicability to many different types of experiments.

@@

Here simulations were generated mimicking a Label-free experiment with a high number of modified and unmodified peptide features and no missing values. Multiple simulations were generated with different numbers of replicates and conditions. In all simulations 1000 modified peptides were generated, half of which were true positives and half true negatives. Half of the true positives and true negatives were generated convoluted with global protein abundance. Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential. Modified peptides and unmodified proteins were both simulated with 10 features.

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed\cite{Zhu}. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a Label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}

@@

\subsubsection*{Proposed approach is less effective without a separate global profiling run but still fixes bias in Label-free experiment}

@@

Additionally, expression levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide. In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified. One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution. However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites. A more complex approach to addressing this problem is most likely necessary.

>>>>

The scientific community increasingly utilizes mass spectrometry (MS)-based proteomics to detect changes in post-translational modifications (PTMs) between samples from various conditions. Experiments targeting PTMs face numerous challenges. These include the low abundance of modified proteoforms, few representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches to estimating relative changes in PTMs, and to detecting differential PTM abundance, should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider relevant sources of confounding and variation. In this manuscript we propose such a general statistical framework, which is comprehensive, accurate, and leads to reproducible results. The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}, \todo{and is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}. The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with \todo{which?} acquisition methods \todo{demonstrate the improved - what? - as compared to the currently used approaches for relative PTM quantification}. The proposed framework is implemented in the free and open-source R package $MSstatsPTM$.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe,Cohen}. Bottom-up liquid chromatography coupled with mass spectrometry (LC-MS) is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) \cite{Kall:2011ub,Roepstorff}. Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}. \todo{The previous sentence is great, but I am wondering whether it can lead the reviewers in a wrong direction, since we only address one of these problems } Some scientific questions,  such as discovery of protein biomarkers of a disease, consider meaningful patterns of changes in PTM abundance alone \cite{Petushkova_2017}. Other questions, e.g. in biological and physiological studies based on very short timescales, require us to distinguish the confounding of quantitative changes in PTM and of overall changes of the protein abundance.

Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \todo{refs for these points?}. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complication the interpretation of the results \cite{Olsen:2013}. \todo{Finally, technological aspects of bottom-up MS experiments, and presence of labeling, e.g. by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.}

\todo{Currently statistical analyses of these experiments are performed with methods such as t-test and Limma methods\cite{Ritchie_15a}. They have drawbacks XXX}

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The proposed framework combines the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, and takes into account various sources of variation and confounding. The framework requires an experimental design\todo{that, for each sample, quantifies the peptides with post-translational modifications, as well as  peptides from the same proteins with no modification sites}. \todo{It is applicable to both label-free and tandem mass tag (TMT)-based acquisitions}. The proposed framework summarizes the abundances of \todo{peptides with and without the modification sites}, and fits a linear mixed effects model that automatically adjusts to the biological and technological aspects of the experimental design. \todo{For questions focusing on PTM, can analyze changes in the modified peptides. For eliminating the confounding, combine...} Finally, changes in the PTM and the protein-level abundances are combined to remove bias due to the confounding of these two sources.

We evaluated the proposed framework on datasets from computer simulations, benchmark controlled mixtures, and biological investigations \todo{expand on the relevance and the comprehensive nature of the datasets - diverse organisms, acquisitions}. The evaluations demonstrated that by appropriately leveraging the information from \todo{which sources}, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated proportion of false positive \todo{what?}. \todo{In particular, accounting for the confounding from protein abundance...}

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014,Huang:2020}. \todo{we claim that the results are reproducible - what does this mean?}

@@

The datasets were chosen to provide a performance benchmark of the proposed approach in situations with known ground truth, and to represent a variety of experimental designs and acquisition methods. The computer simulations varied in experimental realism. The first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in a real experiment with known \todo{changes in what?}. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, experimental designs, and data acquisition strategies.

@@

The simulation mimicked a label-free experiment with a high number of modified and unmodified peptide features and no missing values. Multiple datasets were generated with different numbers of replicates and conditions. In all simulations 1000 modified peptides were generated, half of which were true positives and half true negatives. Half of the true positives and true negatives were generated convoluted with global protein abundance. Before adjustment, the convoluted true positives should appear to be negatives, while the convoluted true negatives would appear differential. Modified peptides and unmodified proteins were both simulated with 10 features.

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential abundance analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed\cite{Zhu}. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

@@

\todo{Check that this section uses past tense; look out for informal expressions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}

@@

\subsubsection*{In label-free experiment without a separate global profiling run, the proposed approach could eliminate the bias \todo{due to what?}, albeit less effectively than \todo{in presence of a global profiling run?}}

@@

Additionally, abundance levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide. In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified. One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution. However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites. A more complex approach to addressing this problem is most likely necessary.
